Meta analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)

### 2017-05-17 code/Rmd/MetaAnalysis.Rmd
Major changes

library(Vennerable)
library(colorfulVennPlot)
library(xlsx)

@@

gsets = c("GO_Biological_Process_2015", "Reactome_2016", "WikiPathways_2016")
GeneSets.Enrichr = GeneSets[gsets]

@@

# Download cell type markers from synapse (list from GSE73721 by Mariet)
load(synGet('syn9818080')@filePath)
all.used.ids = c(all.used.ids, 'syn9818080')

GeneSets = c(list(CellTypeMarkers = CellTypeList,
                AD = AD),
                GeneSets.Enrichr)
```

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
all.used.ids = covariates.ids
covariates = lapply(covariates.ids, downloadFile)

covariates$ROSMAP$BrainRegion = 'DLPFC'
covariates$MSBB = covariates$MSBB %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
covariates$MAYO = covariates$MAYO %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

covariates = covariates %>% 
  rbindlist(use.names = T, fill = T, idcol = 'Study') %>%
  dplyr::select(SampleID, Diagnosis, BrainRegion)

@@

tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05')
tmp %>%
  group_by(Study, BrainRegion, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSSM', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+BrainRegion) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Pathway enrichment analysis (individual brain regions)
```{r pathway.enrich}
# setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
#   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
#   dlply(.(BrainRegion), .fun = function(x){
#     setdiff(unique(x$hgnc_symbol),c(''))
#   })

setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) > log2(1.2)) %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, levels = c(-1,1),
                                   labels = c('-1' = 'DOWN', '1' = 'UP'))) %>%
  dlply(.(BrainRegion, Direction), .fun = function(x){
    setdiff(unique(x$hgnc_symbol),c(''))
  })

# Filter gene list
GeneSets = mapply(function(x,y){
  names(x) = paste(y,names(x),sep = ':')
  return(x)
},GeneSets, names(GeneSets), SIMPLIFY = F) %>%
  Reduce(c,.)
GeneSets = GeneSets[setdiff(seq(1,length(GeneSets)), c(grep('Mus musculus', names(GeneSets))))]

GeneSets = lapply(GeneSets, function(x, backgroundGenes){
  x = intersect(x,backgroundGenes)
}, backgroundGenes)
GeneSets = GeneSets[sapply(GeneSets, length) >= 15 & sapply(GeneSets, length) <= 1000]

enrichResults = plyr::ldply(setsToTest, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, GeneSets, backgroundGenes, 
.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')))

tmp = dplyr::select(enrichResults, BrainRegion, .id, fdr) %>%
  dplyr::filter(fdr <= 0.05) %>%
  tidyr::spread(BrainRegion, fdr)

### Get cell type fraction from synapse
```{r celltype.frac, eval = FALSE}
celltype.fraction.ids = c(ROSMAP = 'syn8124520',
                          MSBB = 'syn8146376',
                          MAYO = 'syn8135250')
celltype.fraction = lapply(celltype.fraction.ids, downloadFile)

cell.frac = mapply(function(cov, frc){
  x = filter(frc, SampleID %in% cov$SampleID) %>%
    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
    tidyr::gather(celltype, fraction, -SampleID) %>%
    left_join(cov)
}, covariates, celltype.fraction, SIMPLIFY = F) %>%
  rbindlist(use.names = T, fill = T) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))

pl = list()

p = ggplot(cell.frac %>% filter(celltype %in% c('neurons', 'microglia', 'astrocytes')), 
           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
p = p + facet_grid(celltype~., scales = 'free_y')
p = p + xlab('Brain Region') + ylab('Fraction')
p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
pl[[1]] = p

p = ggplot(cell.frac %>% filter(celltype %in% c('endothelial', 'oligodendrocytes', 'OPC')), 
           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
p = p + facet_grid(celltype~., scales = 'free_y')
p = p + xlab('Brain Region') + ylab('Fraction')
p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
pl[[2]] = p

multiplot(plotlist = pl, cols = 2)
```

### Get hgnc_symbol mapping from biomaRt
```{r biomart}
# Get background gene ids
backgroundGenes = sapply(diff.exp, function(x) x$hgnc_symbol) %>%
  Reduce(c,.) %>% unique() %>% setdiff(c(''))
```

### Consistently differentially expressed genes (in all 7 brain regions)
```{r diffexp.all, fig.height=16, fig.width=10}
# Find genes that are differentially expressed in each brain region
diff.exp.gl = lapply(diff.exp, function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
    dplyr::select(Region, ensembl_gene_id) %>%
    unique
}) %>%
  rbindlist(use.names = T, fill = T) %>%
  dlply(.(Region), .fun = function(x){setdiff(unique(x$ensembl_gene_id),c(''))})

tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(ensembl_gene_id %in% Reduce(c, diff.exp.gl))

tmp1$Direction = 'NONE'
tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC >= log2(1.2)] = 'UP'
tmp1$Direction[tmp1$adj.P.Val <= 0.05 & tmp1$logFC <= -log2(1.2)] = 'DOWN'

tmp = tmp1 %>%
  group_by(ensembl_gene_id) %>%
  summarise(count = sum(Direction != 'NONE')) %>%
  filter(count == 7) %>%
  left_join(dplyr::select(tmp1, ensembl_gene_id, hgnc_symbol) %>% unique)

writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
                 length(unique(tmp$hgnc_symbol))))
kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))

writeLines('Top six differentially expressed genes are')
tmp.eff.sz = filter(tmp1, hgnc_symbol %in% tmp$hgnc_symbol) %>%
  group_by(hgnc_symbol) %>% 
  summarise(es = sum(logFC, na.rm = T)) %>% 
  top_n(6, abs(es))

pl = list()
p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% c('MEIS3', 'PCSK1', 'ADAMTS2')), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('green', 'red'))
pl[[1]] = p

p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% c('FOXJ1', 'PRELP', 'S100A4')), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey', 'red'))
pl[[2]] = p

multiplot(plotlist = pl, cols = 2)
```

### Meta Analysis
```{r metaanalysis.min.pval}
# Get raw p.values
P.Value = diff.exp %>% 
  rbindlist(use.names = T, fill = T) %>%
  dplyr::select(gene.id, P.Value, Region) %>%
  unique() %>%
  tidyr::spread(Region, P.Value)
P.Value[is.na(P.Value)] = 1
rownames(P.Value) = P.Value$gene.id
P.Value$gene.id = NULL
P.Value = data.matrix(P.Value)

# Perform minimum p-value based meta-analysis
min.pval = P.Value %>% 
  apply(1, min, na.rm = T) %>%
  p.adjust(method = 'bonferroni') %>%
  rownameToFirstColumn('gene.id') %>%
  dplyr::rename(min.pval = DF)

# Perform fisher p-value based meta-analysis
fb.pval = P.Value %>%
  fishersPvalue %>%
  p.adjust(method = 'bonferroni') %>%
  rownameToFirstColumn('gene.id') %>%
  dplyr::rename(fb.pval = DF)
```

```{r }
# Rank prod based analysis
expr = resid.expr[!duplicated(resid.expr$ensembl_gene_id),-c(1:2)]
rownames(expr) = resid.expr$ensembl_gene_id[!duplicated(resid.expr$ensembl_gene_id)]

cl = covariates$Status
names(cl) = covariates$ID
cl[cl == 'NCI'] = 'Control'
cl[cl == 'ND'] = 'Control'
cl[cl == 'SD'] = 'AD'
cl = cl[cl %in% c('Control','AD')] %>% droplevels 
cl.tmp = -as.numeric(cl)+2
names(cl.tmp) = names(cl)

origin = covariates$BrainRegion %>% as.numeric
names(origin) = covariates$ID
origin = origin[names(cl)]

expr = expr[, names(cl)]

rankprod.pfp = RPadvance(expr, cl.tmp, origin, num.perm = 100, gene.names = rownames(expr), plot=T, rand = 123456, huge=T)
rankprod.pval = apply(rankprod.pfp$pfp, 1, min) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  dplyr::rename(rankprod.pval = DF)

metaAnal = join_all(list(min.pval, max.pval, fb.pval, rankprod.pval), type = 'full')
metaAnal$weight = -log10(metaAnal[,-c(1,6)]) %>% rowSums %>% rank
metaAnal = metaAnal %>% arrange(desc(weight))

writeLines('Number of genes differentially expressed between AD and control at an FDR of 0.05 are')
tmp = metaAnal %>% 
  dplyr::select(-weight) %>%
  gather(method, pval, -ensembl_gene_id) %>% 
  group_by(method) %>% 
  filter(pval <= 0.05) %>%
  dplyr::select(ensembl_gene_id) %>%
  unique %>%
  dlply(.(method), .fun = function(x){unique(x$ensembl_gene_id)})
kable(sapply(tmp, length) %>% as.data.frame)

tmp1 = Venn(tmp)

plotVenn3d(tmp1@IndicatorWeight[-(1),'.Weight'], labels = names(tmp))
```
### Heatmap of expression
```{r heatmap}
# Get the overlapping gene sets from all three methods
tmp.expr = t(scale(t(expr)))
tmp.expr = scale(tmp.expr)
tmp.expr = tmp.expr[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),]

tmp.covariates = dplyr::select(covariates, BrainRegion, Status)
tmp.covariates$Status[tmp.covariates$Status == 'NCI'] = 'Control'
tmp.covariates$Status[tmp.covariates$Status == 'ND'] = 'Control'
tmp.covariates$Status[tmp.covariates$Status == 'SD'] = 'AD'
tmp.covariates = tmp.covariates %>% droplevels
rownames(tmp.covariates) = covariates$ID

h1 = HeatmapAnnotation(tmp.covariates)
Heatmap(tmp.expr, top_annotation = h1, show_row_names = F, show_column_names = F,
        show_row_dend = F, show_column_dend = F)
```

@@

# Store covariates in synapse
write.table(covariates, file = 'ALLCovariates.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLCovariates.tsv', 
           name = 'Covariates', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store logcpm in synapse
write.table(logcpm, file = 'ALLLogCPM.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLLogCPM.tsv', 
           name = 'Log CPM (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store residual expression in synapse
write.table(expr, file = 'ALLExpr.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLExpr.tsv', 
           name = 'Residual Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store diffexp in synapse
write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLDiffExp.tsv', 
           name = 'Differential Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store meta analysis in synapse
write.table(metaAnal, file = 'ALLMetaAnal.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLMetaAnal.tsv', 
           name = 'Meta Analysis (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```

>>>>

library(Vennerable)
library(colorfulVennPlot)
library(rmeta)

@@

gsets = c("GO_Biological_Process_2015", "Reactome_2016","KEGG_2016")
GeneSets.Enrichr = GeneSets[gsets]

@@

# Download celltype specific genesets from synapse
load(synGet('syn9818080')@filePath)
all.used.ids = c(all.used.ids, 'syn9818080')

GeneSets = c(GeneSets.Enrichr, 
             list(CellTypeMarkers = CellTypeList,
                  AD = AD))

GeneSets = do.call(c, GeneSets)
```

### Get covariates/sample metadata from synapse
```{r covariates}
covariates = downloadFile('syn9765937')
all.used.ids = c(all.used.ids, 'syn9765937')

@@

tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
tmp %>%
  group_by(Study, BrainRegion, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSSM', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+BrainRegion) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)

diffExpSet = plyr::dlply(tmp, .(BrainRegion), .fun = function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2), adj.P.Val <= 0.05) %>%
    dplyr::select(hgnc_symbol) %>%
    unlist %>%
    as.character() %>%
    unique() %>%
    setdiff('')
})
```

### Meta Analysis
```{r metaanalysis.min.pval}
library(metamisc)
metaAnal <- function(diff.exp, region){
  # Get raw p.values
  P.Value = diff.exp %>% 
    rbindlist(use.names = T, fill = T) %>%
    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
    dplyr::select(ensembl_gene_id, P.Value, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, P.Value)
  P.Value[is.na(P.Value)] = 1
  rownames(P.Value) = P.Value$ensembl_gene_id
  P.Value$ensembl_gene_id = NULL
  P.Value = data.matrix(P.Value)

  # Perform fisher p-value based meta-analysis
  fb.pval = P.Value %>%
    fishersPvalue %>%
    p.adjust(method = 'fdr') %>%
    rownameToFirstColumn('gene.id') %>%
    dplyr::rename(fb.pval = DF)
  sum(fb.pval$fb.pval <= 0.05)

  # Dersimonian based meta analysis
  effect.size = diff.exp %>% 
    rbindlist(use.names = T, fill = T) %>%
    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
    dplyr::select(ensembl_gene_id, logFC, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, logFC)
  rownames(effect.size) = effect.size$ensembl_gene_id
  effect.size$ensembl_gene_id = NULL
  effect.size[is.na(effect.size)] = 0

  effect.var = diff.exp %>% 
    rbindlist(use.names = T, fill = T) %>%
    dplyr::filter(Comparison == 'CONTROL-AD', BrainRegion %in% region) %>%
    dplyr::mutate(sd = (CI.R - logFC)/3,
                  var = sd^2) %>%
    dplyr::select(ensembl_gene_id, var, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, var)
  rownames(effect.var) = effect.var$ensembl_gene_id
  effect.var$ensembl_gene_id = NULL
  effect.var[is.na(effect.var)] = 0

  ds.metaAnal = ldply(1:dim(effect.size)[1], function(i, effect.size, effect.var){
    es = t(effect.size[i,]); es = es[es!=0]
    ev = t(effect.var[i,]); ev = ev[ev!=0]
  fit = uvmeta(es, ev, model = "random", method = "MOM")
  data.frame(ensembl_gene_id = rownames(effect.size)[i],
             logFC = fit$results[1,1],
             CI.L = fit$results[1,1] - 3 * sqrt(fit$results[1,2]),
             CI.R = fit$results[1,1] + 3 * sqrt(fit$results[1,2]))
  }, effect.size, effect.var)

  tmp = dplyr::rename(fb.pval, ensembl_gene_id = gene.id) %>%
    left_join(ds.metaAnal) %>%
    dplyr::filter(fb.pval <= 0.05, sign(CI.L) == sign(CI.R), abs(logFC) >= log2(1.2)) %>%
    dplyr::arrange(desc(logFC)) %>%
    left_join(diff.exp %>%
                rbindlist(use.names = T, fill = T) %>%
                dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
                unique())
}

meta.anal.all= metaAnal(diff.exp, c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
meta.anal.imp= metaAnal(diff.exp, c('DLPFC', 'TCX', 'PHG'))
diffExpSet = c(diffExpSet, list(METANLZ.ALL = setdiff(meta.anal.all$hgnc_symbol, ''),
                                METANLZ.T3 = setdiff(meta.anal.imp$hgnc_symbol, '')))
```

### Pathway enrichment analysis (individual brain regions)
```{r pathway.enrich}
backgroundGenes = diff.exp %>%
  rbindlist(use.names = T, fill = T) %>%
  dplyr::select(hgnc_symbol) %>%
  unlist() %>%
  setdiff('')

# Filter geneset
GeneSets = lapply(GeneSets, function(x, backgroundGenes){
  x = intersect(x,backgroundGenes)
}, backgroundGenes)
GeneSets = GeneSets[sapply(GeneSets, length) >= 15 & sapply(GeneSets, length) <= 1000]

enrichResults = plyr::ldply(diffExpSet, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T, .id = 'Category') %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, GeneSets, backgroundGenes, 
.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')),
.id = 'BrainRegion')

tmp = dplyr::select(enrichResults, BrainRegion, Category, fdr, Odds.Ratio)
tmp = tmp[c(grep('CellTypeMarkers', tmp$Category),
            grep('GeneticLoci', tmp$Category)), ]
tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
tmp = tmp %>%
  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
  tidyr::spread(BrainRegion, Odds.Ratio)
rownames(tmp) = gsub('CellTypeMarkers.','',tmp$Category) %>%
  gsub('_Zhang_Neuron_2015','',.)
tmp$Category = NULL

h = ComplexHeatmap::Heatmap(tmp, name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,4), 'mm'))

tmp = dplyr::select(enrichResults, BrainRegion, Category, fdr, Odds.Ratio)
tmp = tmp[c(grep('Reactome', tmp$Category),
            grep('GO', tmp$Category)), ] %>%
  unique()
tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
tmp = tmp %>%
  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
  tidyr::spread(BrainRegion, Odds.Ratio)
tmp = tmp[rowSums(tmp[,-(1)]) != 0, c(1, which(colSums(tmp[,-(1)]) != 0) + 1)]
rownames(tmp) = gsub('GO_Biological_Process_2015.','',tmp$Category) %>%
  gsub('Reactome_2016.','',.)
tmp$Category = NULL

h = ComplexHeatmap::Heatmap(tmp, name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,4), 'mm'))
```

### Gender specific differential expression (individual brain regions)
```{r gender.specific}
load(synGet('syn9766351')@filePath)

tmp = ldply(all.results, .fun = function(x){
  x$Direction = 'NONE'
  x$Direction[x$adj.P.Val <= 0.05 & x$logFC <= -log2(1.2)] = 'DOWN'
  x$Direction[x$adj.P.Val <= 0.05 & x$logFC >= log2(1.2)] = 'UP'

  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.FEMALE-CONTROL.FEMALE') %>%
    dplyr::group_by(Direction) %>%
    dplyr::summarise(count = n())
}) %>%
  dplyr::mutate(Direction = paste(Direction,'FEMALE',sep = '.')) %>%
  tidyr::spread(Direction, count) %>%
  left_join(
    ldply(all.results, .fun = function(x){
      x$Direction = 'NONE'
      x$Direction[x$adj.P.Val <= 0.05 & x$logFC <= -log2(1.2)] = 'DOWN'
      x$Direction[x$adj.P.Val <= 0.05 & x$logFC >= log2(1.2)] = 'UP'

      dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.MALE-CONTROL.MALE') %>%
        dplyr::group_by(Direction) %>%
        dplyr::summarise(count = n())
    }) %>%
      dplyr::mutate(Direction = paste(Direction,'MALE',sep = '.')) %>%
      tidyr::spread(Direction, count))

metaAnal.gender <- function(allResults, comp, region){
  # Get raw p.values
  P.Value = allResults %>%
    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
    dplyr::select(ensembl_gene_id, P.Value, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, P.Value)
  P.Value[is.na(P.Value)] = 1
  rownames(P.Value) = P.Value$ensembl_gene_id
  P.Value$ensembl_gene_id = NULL
  P.Value = data.matrix(P.Value)

  # Perform fisher p-value based meta-analysis
  fb.pval = P.Value %>%
    fishersPvalue %>%
    p.adjust(method = 'fdr') %>%
    rownameToFirstColumn('gene.id') %>%
    dplyr::rename(fb.pval = DF)
  sum(fb.pval$fb.pval <= 0.05)

  # Dersimonian based meta analysis
  effect.size = allResults %>% 
    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
    dplyr::select(ensembl_gene_id, logFC, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, logFC)
  rownames(effect.size) = effect.size$ensembl_gene_id
  effect.size$ensembl_gene_id = NULL
  effect.size[is.na(effect.size)] = 0

  effect.var = allResults %>% 
    rbindlist(use.names = T, fill = T, idcol = 'BrainRegion') %>%
    dplyr::filter(Comparison == comp, BrainRegion %in% region) %>%
    dplyr::mutate(sd = (CI.R - logFC)/3,
                  var = sd^2) %>%
    dplyr::select(ensembl_gene_id, var, BrainRegion) %>%
    unique() %>%
    tidyr::spread(BrainRegion, var)
  rownames(effect.var) = effect.var$ensembl_gene_id
  effect.var$ensembl_gene_id = NULL
  effect.var[is.na(effect.var)] = 0

  ds.metaAnal = ldply(1:dim(effect.size)[1], function(i, effect.size, effect.var){
    es = t(effect.size[i,]); 
    ev = t(effect.var[i,]); 
    es1 = es[es != 0 & ev != 0]; ev1 = ev[ev != 0 & es != 0]
    fit = uvmeta(es1, ev1, model = "random", method = "MOM")
    data.frame(ensembl_gene_id = rownames(effect.size)[i],
               logFC = fit$results[1,1],
               CI.L = fit$results[1,1] - 3 * sqrt(fit$results[1,2]),
               CI.R = fit$results[1,1] + 3 * sqrt(fit$results[1,2]))
  }, effect.size, effect.var)

  tmp = dplyr::rename(fb.pval, ensembl_gene_id = gene.id) %>%
    left_join(ds.metaAnal) %>%
    dplyr::filter(fb.pval <= 0.05, sign(CI.L) == sign(CI.R)) %>%
    dplyr::arrange(desc(logFC)) %>%
    left_join(diff.exp %>%
                rbindlist(use.names = T, fill = T) %>%
                dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
                unique())
}

meta.female = metaAnal.gender(all.results, comp = 'AD.FEMALE-CONTROL.FEMALE', 
                              region = c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
meta.male = metaAnal.gender(all.results, comp = 'AD.MALE-CONTROL.MALE',
                            region = c('DLPFC', 'CER', 'TCX', 'FP', 'STG', 'PHG', 'IFG'))
meta.female.imp = metaAnal.gender(all.results, comp = 'AD.FEMALE-CONTROL.FEMALE', 
                                  region = c('DLPFC', 'TCX', 'PHG'))
meta.male.imp = metaAnal.gender(all.results, comp = 'AD.MALE-CONTROL.MALE',
                                region = c('DLPFC', 'TCX', 'PHG'))

diffExpSet.gender = llply(all.results, .fun = function(x){
  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.FEMALE-CONTROL.FEMALE') %>%
    dplyr::left_join(diff.exp %>%
                       rbindlist(use.names = T, fill = T) %>%
                       dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
                       unique) %>%
    dplyr::select(hgnc_symbol) %>%
    unlist() %>% setdiff('')
})

diffExpSet.gender = c(diffExpSet.gender, llply(all.results, .fun = function(x){
  dplyr::filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), Comparison == 'AD.MALE-CONTROL.MALE') %>%
    dplyr::left_join(diff.exp %>%
                       rbindlist(use.names = T, fill = T) %>%
                       dplyr::select(ensembl_gene_id, hgnc_symbol) %>%
                       unique) %>%
    dplyr::select(hgnc_symbol) %>%
    unlist() %>% setdiff('')
}))
names(diffExpSet.gender)[1:7] = paste('FEMALE',names(diffExpSet.gender)[1:7],sep = '.')
names(diffExpSet.gender)[8:14] = paste('MALE',names(diffExpSet.gender)[8:14],sep = '.')
diffExpSet.gender$FEMALE.METANALZ.ALL = setdiff(meta.female$hgnc_symbol,'')
diffExpSet.gender$MALE.METANALZ.ALL = setdiff(meta.male$hgnc_symbol,'')
diffExpSet.gender$FEMALE.METANALZ.T3 = setdiff(meta.female.imp$hgnc_symbol,'')
diffExpSet.gender$MALE.METANALZ.T3 = setdiff(meta.male.imp$hgnc_symbol,'')

enrichResults.gender = plyr::ldply(diffExpSet.gender, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T, .id = 'Category') %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, GeneSets, backgroundGenes, 
.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')),
.id = 'BrainRegion')

tmp = dplyr::select(enrichResults.gender, BrainRegion, Category, fdr, Odds.Ratio)
tmp = tmp[c(grep('CellTypeMarkers', tmp$Category),
            grep('GeneticLoci', tmp$Category)), ]
tmp$Odds.Ratio[tmp$fdr > 0.05] = 0
tmp = tmp %>%
  dplyr::select(BrainRegion, Category, Odds.Ratio) %>%
  tidyr::spread(BrainRegion, Odds.Ratio)
rownames(tmp) = gsub('CellTypeMarkers.','',tmp$Category) %>%
  gsub('_Zhang_Neuron_2015','',.)
tmp$Category = NULL

h = ComplexHeatmap::Heatmap(t(tmp), name = 'Odds Ratio', col = wesanderson::wes_palette('Zissou', 4))
draw(h, heatmap_legend_side = 'left', padding = unit(c(4,2,2,10), 'mm'))

```

@@
